Case-based Arabic Morphological Analysis

### 2010-12-16 trunk/tex/rp.tex
refinements+sort bib

Most of academic research in the field of Arabic text mining 
is covered in the almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}.
The book summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest in this area with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While a non-Arabic user is interested in machine translation to 
help him understand an Arabic document, 
an Arabic user is more interested in 
information hard to discover manually such as 
the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness. 
Another Arabic user may be interested in looking at the relation
between stock prices and political news or the 
relation between suspect profiles in interrogation reports 
and investigative reports from the crime scenes.

We propose to build a 
%{\em relational text mining 
%of Arabic using morphological and case-based analysis}. 
{\em case-based Arabic text analysis using relations and morphology}(CATARM) framework. 
CATARM will take as input sets of documents, each of a specific
type, and provides the user with a visual query language that
allows him to build a flow of Arabic text mining
modules and discover desired relations between documents. 
CATARM will translate the query into a netlist of computational 
components, it will also extract parameters from the query and 
the document sets and use them to simplify the computational 
components. 
%The main CATRAM computational components will be finite 
%state machines (FSM) that express a linguistic model of the Arabic 
%language and that can be parametrized and relaxed to fit a case 
%study.

@@

%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{almdrsT} ends with
\RL{T} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

Current morphological analyzers such as 
Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
These analyzers are used in several open source spell checkers as 
well as NLP frameworks~\cite{Col09}.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

{\bf Hypothesis 1.} We hypothesize that many case studies, 
in particular relational
case studies, do not require high accuracy from the 
low level morphological analysis.
The sophistication of the sought higher level query can eliminate 
many of the ambiguous low level solutions at a lower computational 
cost.
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we do not need to further analyze the rest of the word 
to provide an answer.

We find evidence to our hypothesis in~\cite{Maamouri:10}. The 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
required the high NLP task to interact with the low 
level morphological analyzer.
This led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our preliminary
results that a 
case-based morphological analyzer that allows a high level 
case-based controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

%Prior work in Arabic text mining addressed the classification and 
%categorization of Arabic text and made little use of features 
%unique to the Arabic language.
Researchers in~\cite{AEL07,Ham07,Abd07,MEl03} 
preprocessed the Arabic text to make it presentable to known text 
mining techniques that tend to Latin text and queries.
%Buckwalter~\cite{Bee89,Tim04} uses morphological analysis to enable 
%relaxed stemming of Arabic words.
El-Halees used statistical analysis to address the problem of 
classifying Arabic text documents with maximum 
entropy~\cite{AEL07}.
In~\cite{Abd07} Mesleh evaluated six different techniques to 
classify Arabic text using a support vector machine.
El Dost~\cite{MEl03} used root words in the Arabic language to 
speed up automated and hierarchical indexing.
Al-Zoghby~\cite{Ham07}
%addresses the problem of 
%entities extracted from Arabic text with similar content but 
%slightly different names via 
used the derivation feature in 
the Arabic language as a similarity measurement.
This work is limited to the names of the fields in the data 
structures extracted from Arabic texts.

@@

%\subsection{\bf Arabic text mining in industry}
Few companies in the industry provide primitive Arabic text 
analysis tools.
Sakhr~\cite{Sak09} provides a categorization 
tool with a limited number of predefined categories, and a summary 
generator that selects sentences from text.
Basis technology~\cite{Bas09} provides Rosette, a lexical 
analysis tool that normalizes Arabic text as a pre-processing step 
for indexers and other general purpose data mining 
tools.  Basis also provides REX; an entity extractor.

@@

{\bf Goals.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.

@@

In recent research efforts, the CO-LPI is investigating 
keyphrase extraction fro Arabic documents.

The Co-Lead PI published extensively in this domain. 
A selected list of her recent publications follows. 
\begin{itemize}\itemsep0pt
\item Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, 
"Feature Reduction Techniques for Arabic Text Categorization". 
Journal of the American Society for Information Science and Technology 
(JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
\item Rehab M. Duwairi, "Arabic Text Categorization", 
International Arab Journal for Information Technology (IAJIT), 
Volume 4, Number 2, pages 125  132, 2007.
\item Rehab M. Duwairi, "Machine Learning for Arabic Text Categorization". 
Journal of the American Society for Information Science and Technology 
(JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
\end{itemize}

The Lead PI worked in the area of software arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of literature heritage~\cite{Zaraket06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

The lead PI holds several approved patents and has several 
patents pending approval in the area of logic verification 
and synthesis.
The lead PI worked and published about relational logic and 
logic solvers applied to structured languages as follows.
\begin{itemize}
\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
\end{itemize}

>>>>

The almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}
covers most academic research work on Arabic text mining.
It summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While machine translation is of gneral interest, 
an Arabic user is more interested in 
information hard to discover manually.
Examples are {the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness,
the relation
between stock prices and political news, or the 
relation between interrogation reports with suspects
and investigative reports from the crime scenes.

We propose to build a framework for
%{\em relational text mining 
%of Arabic using morphological and case-based analysis}. 
{\em case-based Arabic text analysis using relations and morphology}(CATARM). 
CATARM takes as input several sets of documents, each of a specific
type, and a user query. % expressed in a visual language.
The query, expressed in a visual language, allows the user 
to build a flow of Arabic text mining
components and discover desired relations between documents. 
CATARM compiles the query into a netlist of computational 
components. It also uses features extracted from the query and 
the document sets to simplify the computational 
components. 
%The main CATRAM computational components will be finite 
%state machines (FSM) that express a linguistic model of the Arabic 
%language and that can be parametrized and relaxed to fit a case 
%study.

\novocalize
%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

@@

Current morphological analyzers such as 
Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist
and are used in several open source spell checkers as 
well as NLP frameworks~\cite{Col09}.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

{\bf Hypothesis 1.} We hypothesize that many case studies, 
in particular relational ones, 
do not require high accuracy from the 
low level morphological analysis.
The sophistication of the higher level query can eliminate 
many ambiguous low level solutions at a lower computational 
cost.
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we no further need to analyze the rest of the word.

We find evidence to our hypothesis in~\cite{Maamouri:10} where
a new corpus was added to the Arabic Tree Bank~\cite{Maamouri:04}.
The addition requires compliance with 
the DARPA GALE program including several NLP tasks such as 
English translation annotations and word-level alignment
of Arabic and English. 
The new genre of data challenged the existing data processing 
and annotation guidelines. 
Alternative guidelines came to place where the 
high level NLP task interacted with the 
morphological analyzer.% along with other concurrent tasks. 
This led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
For example, the analyzers may find different solutions, and may 
present the same solutions in different orders.
We find strong evidence in our preliminary
results that a 
case-based morphological analyzer that allows a 
case-based controller to intervene at every decision to 
guide, use, and refine the morphological analysis
is of high utility.

%Prior work in Arabic text mining addressed the classification and 
%categorization of Arabic text and made little use of features 
%unique to the Arabic language.
Researchers in~\cite{AEL07,Ham07,Abd07,MEl03} 
preprocessed the Arabic text and presented it to known 
techniques that tend to Latin text and queries.
%Buckwalter~\cite{Bee89,Tim04} uses morphological analysis to enable 
%relaxed stemming of Arabic words.
El-Halees used statistical analysis to address the problem of 
classifying Arabic text documents with maximum 
entropy~\cite{AEL07}.
In~\cite{Abd07} Mesleh evaluated six different techniques to 
classify Arabic text using a support vector machine.
El Dost~\cite{MEl03} used root words in the Arabic language to 
speed up automated and hierarchical indexing.
Al-Zoghby~\cite{Ham07}
%addresses the problem of 
%entities extracted from Arabic text with similar content but 
%slightly different names via 
used the derivation feature in 
the Arabic language as a similarity measurement.
%This work is limited to the names of the fields in the data 
%structures extracted from Arabic texts.

@@

%\subsection{\bf Arabic text mining in industry}
Few companies in the industry provide Arabic text 
analysis tools.
Sakhr~\cite{Sak09} provides a categorization 
tool with a limited number of predefined categories.
Sakhr also provides a summary 
generator that selects important sentences from text.
Basis technology~\cite{Bas09} provides Rosette, a lexical 
analysis tool that normalizes Arabic text as a pre-processing step 
for general purpose data mining 
tools.  Basis also provides REX; an entity extractor.

@@

{\bf Goals.} CATARM 
tends to relational Arabic text mining using morphological analysis.
{\bf (1.)} CATARM will build a novel parametrized 
linguistic computational model (LCM) of Arabic. 
The LCM will be implemented using efficient structures. 
 {\bf (2.)} The LCM will be amenable to Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the relational user queries.
{\bf (3.)} The LCM will be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as 
hidden Markov models (HMM)  and 
latent semantic indexing~\cite{LSI89}.
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from text 
documents and uses set arithmetic and graph analysis techniques
to establish relations between the graphs. {\bf (5.)} CATARM will 
solve at least three interesting case studies such as those
presented in Sections~\ref{s:design:lit} and~\ref{s:design:sec}.

@@

The Co-Lead PI published extensively in the 
domain~\cite{Duw06, Duw07, Duw09}.
In recent research efforts, the CO-LPI is investigating 
keyphrase extraction fro Arabic documents.

The Lead PI worked in the area of software arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of literature heritage~\cite{Zar06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

The lead PI holds several approved patents and has several 
patents pending approval in the area of logic verification 
and synthesis.
The lead PI worked and published about relational logic and 
logic solvers applied to structured languages~\cite{seraICSE07,Zara05}
